---
title: "Reproducible Research: Peer Assessment 1"
author: K. Yang
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(cache = TRUE)
```

## Loading and preprocessing the data
```{r load}
file <- read.csv("activity.csv")
df <- data.frame(file)
#head(df)
#str(df)
df$date <- as.Date(df$date, "%Y-%m-%d")

# Not every interval is recorded. 
# length(unique(df$interval))=288
# interval:0 -- 2355.
```



## What is mean total number of steps taken per day?

```{r mean}

library(dplyr)

dailySteps <- df %>%
        group_by(date) %>%
        summarise(n=sum(steps, na.rm=TRUE))

hist(dailySteps$n, breaks = 20, main="Number of Steps taken daily", xlab="") 
mean <- round(mean(dailySteps$n, na.rm=TRUE), 0.1)
median <- median(dailySteps$n, na.rm=TRUE)
```

**Summary**  
The mean of the total number of steps taken per day: `r mean`  
The median of the total number of steps taken per day: `r median`



## What is the average daily activity pattern?
```{r averageDailyPattern}
library(ggplot2)

pattern <- data.frame(interval=unique(df$interval))

dailyAverage <- with(df, tapply(steps, as.factor(interval), mean, na.rm=TRUE)) # calculate a mean and send to all applicable rows. 

pattern$dailyAverage <- as.integer(round(dailyAverage, 0.1)) # round up to an integer.

# geom_line() requires x values are integers, not factors. 
g <- ggplot(pattern, aes(x=interval, y=dailyAverage)) +
        geom_line() +
        labs (x="Interval", y="Average daily steps")

print(g)

max <- pattern$interval[which.max(pattern$dailyAverage)]

```


**Report**

The interval that contains the maximum number of steps is `r max`.


## Imputing missing values

```{r}
## check each row to identify missing steps.
identifyStatus <- function(record){
        if(is.na(record[1])) return("Missing")
        else return("Complete")
}

status <- apply(df, 1, identifyStatus)
numOfMissing <- length(which(status=="Missing"))

## Fill in the missing data(steps) by assigning the average steps in that given interval.

df2 <- df  # copy the original data frame.

for(i in 1:nrow(df2)){
        if(is.na(df2$steps[i])) {
                interval <- df2$interval[i]
                df2$steps[i] <- pattern$dailyAverage[which(pattern$interval==interval)]
        } else {
                next
        }
}

dailySteps2 <- df2 %>%
        group_by(date) %>%
        summarise(n=sum(steps))

hist(dailySteps2$n, breaks = 20, main="Number of Steps taken daily", xlab="") 
mean2 <- as.integer(mean(dailySteps2$n))
median2 <- median(dailySteps2$n) 


```
**Report**

1. The total number of rows with NA: `r numOfMissing`

2. After filling in the missing data:  
The mean of the total number of steps taken per day: `r mean2`  
The median of the total number of steps taken per day: `r median2`

3. Both mean and median are greater after imputing missing data than before imputing missing data. 



## Are there differences in activity patterns between weekdays and weekends?

```{r}
dayFactor <- factor(c("weekday", "weekend"))

setDay <- function(dateObj){
        day <- weekdays(dateObj)
        if(day=="Saturday" | day=="Sunday") return(dayFactor[2])
        else return(dayFactor[1])
}


df2$day <- sapply(df2$date, setDay)

pattern2 <- df2 %>%
  group_by(interval, day) %>%
  summarise(dailyAverage=mean(steps))

q <- ggplot(pattern2, aes(x=as.integer(interval), y=dailyAverage)) +
    facet_grid(day~.) +   
    geom_line() +
    labs (x="Interval", y="Average daily steps")

print(q)


```

